---
title: "GEDI ratio cutoff analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GB-JL-1189
### Researchers: Jeremy Linsley
### Staff members: Reuben Thomas AND Min-Gyoung Shin


### Dataset:


Neuronal GEDI ratio values indicating live and dead cell status

### Questions:

GEDI is designed to ratio-metrically increase in signal when a cell dies. Because of the kinetics of the indicator (which I can describe in more detail later), dead cells have a much larger and non-linearly increased signal compared to live cells. This results in two major populations of neurons (see dot plots/density plots in figure B and F below).

In the paper, we suggest to distinguish live from dead that we kill a well of cells to empirically identify the dead signal in comparison to the live signal. But sometimes that is not possible, either because cells are precious, or as we’re finding out recently, some cell types are pretty resistant to being killed! So I was wondering if we can model data we already have to generate a predictive formula for what constitutes a dead cell without having to empirically determine a threshold for dead signal. In a lot of the data we have, there are two obvious populations of live and dead, but we are seeing that a lot of the time, there is an obvious live population, but there is not enough death present to find a cluster of dead cells (see second plot). When you look at individual cells with high ratios, its clear that they are indeed dead, but they are just so rare it is hard to generalize the separation of their signal. 

I was thinking we could model this by assuming a normal distribution of live cells, and establishing rules for where the dead cells should be. We will likely need to adjust for variability in imaging (the ratios are not absolute, but relative to the intestines of the two channels), but it seems like this should be doable.


### Analysis Plan 

- Visualization of the distribution of GEDI ratios across time points
- Supervised Gaussian mixture model testing to test annotated GEDI ratios and verify accuracy of prediction
- Define outliers corresponding to dead cells using the Hampel filter and Rosner's test.

### Results

#### Visualization of the distribution of GEDI ratios across time points
```{r}


setwd("~/Dropbox (Gladstone)/GB-JL-1189-Jeremy-GEDI")

library(ggplot2)
MN1<-read.csv("TimelapseDataset1_MNs.csv")
MN1$Timepoint=as.factor(MN1$Timepoint)

bp <- ggplot(MN1, aes(x=Timepoint, y=GEDIratio, fill=Timepoint)) + 
  geom_violin(trim=FALSE) + scale_fill_brewer(palette="Blues") + theme_classic()

print("GEDI ratio distribution per time point (MN1)")
print(bp)






MN2<-read.csv("TimelapseDataset2_MNs.csv")
MN2$Timepoint=as.factor(MN2$Timepoint)





bp <- ggplot(MN2, aes(x=Timepoint, y=GEDIratio, fill=Timepoint)) + 
  geom_violin(trim=FALSE)+ scale_fill_brewer(palette="Blues") + theme_classic()

print("GEDI ratio distribution per time point (MN2)")
print(bp)





```

#### Supervised Gaussian mixture model testing to test annotated GEDI ratios and verify accuracy of prediction
```{r}
library(mclust)

annotated<-read.csv("/Users/mingyoungshin/Dropbox (Gladstone)/GB-JL-1189-Jeremy-GEDI/Curation_and_GEDI.csv")

#set random seed
set.seed(123)

#take 2/3 of the data as training data
train <- sample(1:nrow(annotated), size = round(nrow(annotated)*2/3), replace = FALSE)
annotated.train <- annotated[train,'GEDIratio']

#take the mean of the votes as the consensus annotation
Class.train <- annotated[train,c("Curator1_Accurate","Curator2_Accurate","Curator3_Accurate","Curator4_Accurate")]
Class.train[is.na(Class.train)]=0.5
Class.train=round(rowMeans(Class.train))

#define the test data
annotated.test <- annotated[-train,'GEDIratio']

#retrieve the annotation of the test data
Class.test <- annotated[-train,c("Curator1_Accurate","Curator2_Accurate","Curator3_Accurate","Curator4_Accurate")]
Class.test=round(rowMeans(Class.test))

#train the model using the train data 
mod1 <- MclustDA(annotated.train, as.vector((Class.train)), modelType = "EDDA")

#perform cross validation 
cv <- cvMclustDA(mod1)
```

##### Prediction error in supervised Mclust Gaussian mixed model based on annotated data

```{r}
#retrieve the prediction error
unlist(cv[c('error')])
```



#### Define outliers corresponding to dead cells using the Hampel filter and Rosner’s test 


##### MN1 GEDI ratio distribution and estimated dead cell cutoff
```{r}
library(EnvStats)
library(DescTools)
MN1<-read.csv("TimelapseDataset1_MNs.csv")


options(warn=-1)
upper_bound <- median(sqrt(MN1$GEDIratio)) + 3 * mad(sqrt(MN1$GEDIratio))
upper_bound

hist(sqrt(MN1$GEDIratio),breaks=1000, main=NULL)
#abline(v=upper_bound, col="blue")

outlierC=sum(sqrt(MN1$GEDIratio)>upper_bound)
outlierC


test <- rosnerTest(sqrt(MN1$GEDIratio),
                   k = outlierC)

  outliers=test$all.stats$Value[test$all.stats$Outlier]
  cutoff=min( outliers[which(outliers>median(sqrt(MN1$GEDIratio)))] )
  

abline(v=cutoff,col="red")

mtext(paste0("MN1 combined :cutoff at ",cutoff),cex=1.2)

```

##### MN1 GEDI ratio distribution and estimated dead cell cutoff per time point
```{r}

for(t in 1:max(as.numeric(MN1$Timepoint))){
 
  options(warn=-1)
  
  par(new=FALSE)
  print(t)
  upper_bound <- median(sqrt(subset(MN1,Timepoint==t)$GEDIratio)) + 3 * mad(sqrt(subset(MN1,Timepoint==t)$GEDIratio))
  upper_bound
  
  hist(sqrt(subset(MN1,Timepoint==t)$GEDIratio),breaks=1000,main=NULL)
#  abline(v=upper_bound, col="blue")
  
  outlierC=sum(sqrt(subset(MN1,Timepoint==t)$GEDIratio)>upper_bound)
  outlierC
  
  test <- rosnerTest(sqrt(subset(MN1,Timepoint==t)$GEDIratio),
                     k = outlierC)
  outliers=test$all.stats$Value[test$all.stats$Outlier]
  cutoff=min( outliers[which(outliers>median(sqrt(subset(MN1,Timepoint==t)$GEDIratio)))] )
  
  
  abline(v=cutoff,col="red")
  mtext(paste0("MN1 Time ",t," :cutoff at ",cutoff),cex=1.2)
  
}
```

##### MN2 GEDI ratio distribution and estimated dead cell cutoff
```{r}

options(warn=-1)

MN2<-read.csv("TimelapseDataset2_MNs.csv")

library(EnvStats)
library(DescTools)
### Find Hampel filter: outliers are the values outside the interval the median plus 3 median absolute deviations
upper_bound <- median(sqrt(MN2$GEDIratio)) + 3 * mad(sqrt(MN2$GEDIratio))
upper_bound

hist(sqrt(MN2$GEDIratio),breaks=1000,main=NULL)
#abline(v=upper_bound, col="blue")

### the number of estimated outliers 
outlierC=sum(sqrt(MN2$GEDIratio)>upper_bound)
outlierC

### Use Rosner’s test to find outliers with the given number of estimated number of outliers. Take a square root of the raw value to make the distribution
# of live cells close to normal distribution

test <- rosnerTest(sqrt(MN2$GEDIratio),
                   k = outlierC)

### Find the smallest GEDI value of outlier cells
  outliers=test$all.stats$Value[test$all.stats$Outlier]
  cutoff=min( outliers[which(outliers>median(sqrt(MN2$GEDIratio)))] )

abline(v=cutoff,col="red")
mtext(paste0("MN2 combined :cutoff at ",cutoff),cex=1.2)
```

##### MN2 GEDI ratio distribution and estimated dead cell cutoff per time point
```{r}

for(t in 1:max(as.numeric(MN2$Timepoint))){
  options(warn=-1)
  par(new=FALSE)
  
  upper_bound <- median(sqrt(subset(MN2,Timepoint==t)$GEDIratio)) + 3 * mad(sqrt(subset(MN2,Timepoint==t)$GEDIratio))
  upper_bound
  
  hist(sqrt(subset(MN2,Timepoint==t)$GEDIratio),breaks=1000,main=NULL)
 # abline(v=upper_bound, col="blue")
  
  outlierC=sum(sqrt(subset(MN2,Timepoint==t)$GEDIratio)>upper_bound)
  outlierC
  

  test <- rosnerTest(sqrt(subset(MN2,Timepoint==t)$GEDIratio),
                     k = outlierC)
  outliers=test$all.stats$Value[test$all.stats$Outlier]
  cutoff=min( outliers[which(outliers>median(sqrt(subset(MN2,Timepoint==t)$GEDIratio)))] )
 
  abline(v=cutoff,col="red")
  mtext(paste0("MN2 Time ",t," :cutoff at ",cutoff),cex=1.2)
  
}

```
